---
title: "Monitoring Vocalizations of Threatened Monkey Species in Ecuadorian Secondary Tropical Forests"
subtitle: "Data Analysis"
author: ""
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: yes
    number_sections: yes
    highlight: tango
  html_document:
    theme: readable
    toc: yes
    number_sections: yes
    toc_float: yes
    highlight: tango
header-includes:
  - \renewcommand{\contentsname}{Table of Contents}
encoding: UTF-8
---

# Settings

```{r Markdown setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      fig.align = "center",
                      fig.width = 6,
                      fig.height = 4.5,
                      message = FALSE,
                      warning = FALSE,
                      cache = TRUE)
```

```{r, include = FALSE}
# ggplot settings
palette_HSM <- c(
  "#17283B",
  "#233B57",
  "#2A5779",
  "#3880A9", 
  "#5093A3",  
  "#70A49F",
  "#8FB7A5",
  "#ABC9AD",
  "#D3E1CC",
  "#EDF1E4"
)
```

```{r Libraries import, include = FALSE, message = FALSE}
library(readxl) # Used for reading Excel files (.xls and .xlsx) for data import
library(openxlsx) # To save dataframes into .xlsx format
library(sf) # To import shapefiles
library(elevatr) # Extract elevation data
library(dplyr) # For data manipulation & visualization
library(tidyr) # To reshape, tidy, and organize datasets
library(tibble) # To manipulate tibbles
library(terra) # For handling spatial data
library(stringr) # To manipulate characters
library(ggplot2) # For creating plots
library(rnaturalearth) # To plot maps
library(scales)  # for logarithmic scales
library(corrplot) # To create corrplots
library(GGally) # To investigate correlations
library(mgcv) # To compute GAM
library(ggeffects) # To plot model prediction
library(ggspatial) # To plot spatial data
library(pROC) # To display ROC curves
```

```{r, include = FALSE}
# Define the species names
species_list <- c("Ecuadorian mantled howler monkey", "Brown-headed Spider Monkey")
```

```{r, include = FALSE}
# Raw observations import
df_observations <- read.xlsx("MKY_Data.xlsx", sheet = "observations_filtered") %>%
  filter(is.na(Check)==TRUE) %>% # Remove lines that were manually checked with no detection found
  select(-Check) # Remove check column
```

```{r, include = FALSE}
# Add the Plot_ID, the Soundbox_ID and the Jocotoco_ID for future merge with df_plots
df_observations <- df_observations %>%
  mutate(
    Soundbox_ID = if_else(
      str_detect(File, "/Soundboxes/Plot_\\d+"),
      str_extract(File, "Plot_(\\d+)") %>% str_replace("Plot_", ""),
      NA_character_
    ),
    Plot_ID = str_extract(File, "/data/42-julia-hpc-bio-zoo3/sok73hu/Ecuador/2022/Ecuador2022/Ecuador2022_([^/]+)") %>%
      str_replace("/data/42-julia-hpc-bio-zoo3/sok73hu/Ecuador/2022/Ecuador2022/Ecuador2022_", "") %>%
      str_replace_all("-", ""),  # remove dashes
    Jocotoco_ID = str_match(File, "/CANANDE/\\d{4}-\\d{2}_(\\d{4})-\\d{2}/([^/]+)")[, 2:3] %>%
      apply(1, function(x) {
        year <- x[1]               # second year captured here
        id <- str_remove(x[2], "^C_")  # remove leading 'C_'
        paste0(id, "_", year)
      })
  ) %>%
  mutate(Soundbox_ID = as.numeric(Soundbox_ID)) %>%
  mutate(Jocotoco_ID = if_else(Jocotoco_ID == "NA_NA", NA_character_, Jocotoco_ID))
```

```{r, include = FALSE}
# Description of study sites import
df_plots <- read.xlsx("MKY_Data.xlsx", sheet = "plots_description") %>%
    filter(!duplicated(Jocotoco_ID) | is.na(Jocotoco_ID)) %>%
    filter(!Jocotoco_ID %in% c("G04_2025", "G22_2025", "G03_2025", "G13_2025", "GW62_2025", "G14_2025", "G05_2025", "G28_2025")) # These plots were not analyzed
```

```{r, include = FALSE}
# Rescaling forest cover
df_plots <- df_plots %>%
  mutate(across(
    c(Forest_JRC_500m, Forest_JRC_2000m, Forest_JRCxESA_500m, Forest_JRCxESA_2000m),
    ~ . / 100
  ))
```

```{r, include = FALSE}
# Grid import
df_grid <- read.xlsx("MKY_Data.xlsx", sheet = "grid")

sf_grid <- st_as_sf(
  df_grid,
  coords = c("Longitude", "Latitude"),  # order matters: X (lon) first, Y (lat) second
  crs = 4326)  # WGS84 coordinate reference system

plot(sf_grid, max.plot = 14)
```

```{r, include = FALSE}
# Create a occurence/plot table
df_observations <- df_observations %>%
  select(-Confidence, -Start, -End) %>%
  distinct() %>% # Keep only 1 occurence by audio file
  group_by(Soundbox_ID, Plot_ID, Jocotoco_ID, Common_name) %>%
  summarise(Counts = n(), .groups = "drop")
```

```{r, include = FALSE}
# Duplicate rows for each species
df_data <- df_plots %>%
  mutate(dummy = 1) %>%  # temporary column for cross join
  crossing(Common_name = species_list) %>%
  select(-dummy) %>%  # remove helper column
  mutate(
    Plot_ID = as.character(Plot_ID),
    Soundbox_ID = as.character(Soundbox_ID),
    Jocotoco_ID = as.character(Jocotoco_ID),
    Common_name = as.character(Common_name)
  )
  
# Observations by Plot
obs_by_plot <- df_observations %>%
  filter(!is.na(Plot_ID) & is.na(Soundbox_ID) & is.na(Jocotoco_ID)) %>%
  select(-Soundbox_ID, -Jocotoco_ID) %>%
  mutate(
    Plot_ID = as.character(Plot_ID),
    Common_name = as.character(Common_name)
  )

# Observations by Soundbox
obs_by_soundbox <- df_observations %>%
  filter(!is.na(Soundbox_ID) & is.na(Plot_ID) & is.na(Jocotoco_ID)) %>%
  select(-Plot_ID, -Jocotoco_ID) %>%
  mutate(
    Soundbox_ID = as.character(Soundbox_ID),
    Common_name = as.character(Common_name)
  )

# Observations by Jocotoco_ID
obs_by_jocotoco <- df_observations %>%
  filter(!is.na(Jocotoco_ID) & is.na(Plot_ID) & is.na(Soundbox_ID)) %>%
  select(-Plot_ID, -Soundbox_ID) %>%
  mutate(
    Jocotoco_ID = as.character(Jocotoco_ID),
    Common_name = as.character(Common_name)
  )

# Join to create a complete dataset
df_data <- df_data %>%
  left_join(obs_by_plot, by = c("Plot_ID", "Common_name")) %>%
  left_join(obs_by_soundbox, by = c("Soundbox_ID", "Common_name")) %>%
  left_join(obs_by_jocotoco, by = c("Jocotoco_ID", "Common_name")) %>%
  mutate(
    Counts = coalesce(Counts.x, Counts.y, Counts),
  ) %>%
  select(-Counts.x, -Counts.y)

# Apply 0 to all NAs
df_data <- df_data %>%
  mutate(Counts = if_else(is.na(Counts), 0, Counts)) %>%
  filter(!Common_name=="Colombian white-faced capuchin")

# Have only one Plot_ID
df_data <- df_data %>%
  mutate(Plot_ID = coalesce(Plot_ID, Jocotoco_ID))

# Keep only columns of interest
df_data <- df_data %>%
  select(Plot_ID, Latitude, Longitude, Elevation, Slope, NDVI_500m, NDVI_2000m, Integrity_500m, Integrity_2000m, Forest_ESA_500m, Forest_ESA_2000m, Forest_JRC_500m, Forest_JRC_2000m, D_edge_ESA, D_edge_JRC, D_water, Canopy_height_lang_100m, Canopy_height_tolan_100m, Road_density_500m, Road_density_2000m, Common_name, Counts)

rm(obs_by_plot, obs_by_soundbox, obs_by_jocotoco)
```

```{r, include = FALSE}
# Create a presence/absence column
df_data$Occurrence <- as.numeric(df_data$Counts > 0)
```

```{r}
# Scale Longitude and Latitude for GAM
df_grid <- df_grid %>%
  mutate(Latitude_scaled = scale(Latitude)) %>%
  mutate(Longitude_scaled = scale(Longitude))

df_data <- df_data %>%
  mutate(Latitude_scaled = scale(Latitude)) %>%
  mutate(Longitude_scaled = scale(Longitude))
```

# Resolve pseudo-replication

```{r, include = FALSE}
# Separate spider and howler monkeys
df_data_spider_monkey <- df_data %>%
  filter(Common_name == "Brown-headed Spider Monkey")
df_data_howler_monkey <- df_data %>%
  filter(Common_name == "Ecuadorian mantled howler monkey")
```

We first remove truly duplicated sites in the dataset.

```{r}
# Join detections to grid cells, then deduplicate
deduplicate_by_grid <- function(df_data_species) {
  # Convert species detections to sf object
  sf_data_species <- st_as_sf(df_data_species, 
                     coords = c("Longitude", "Latitude"), 
                     crs = 4326, 
                     remove = FALSE)
  # Subset sf_grid to Grid_ID and Geometry
  sf_grid_subset <- sf_grid %>%
    select(Grid_ID, geometry)
  
  # Spatial join: assign ARU detections to a grid cell
  df_joined <- st_join(sf_data_species, sf_grid_subset, join = st_nearest_feature) %>%
  st_drop_geometry()
  print(df_joined)
  
  # Keep only 1 detection per grid cell and the one with the highest Counts
  df_dedup <- df_joined %>%
    group_by(Grid_ID) %>%
    slice_max(order_by = Counts, n = 1, with_ties = FALSE) %>%
    ungroup()
  
  return(df_dedup)
}

# Apply to both species
df_data_spider_monkey_dedup <- deduplicate_by_grid(df_data_spider_monkey)
df_data_howler_monkey_dedup <- deduplicate_by_grid(df_data_howler_monkey)
```

```{r}
deduplicate_by_distance_v2 <- function(df_data_species, d = 500) {
  
  d <- units::set_units(d, "m")
  
  # Split into zeros and non-zeros
  df_zeros    <- df_data_species %>% filter(Counts == 0)
  df_nonzeros <- df_data_species %>% filter(Counts > 0)
  
  # Convert non-zeros to sf and project
  sf_nonzeros <- st_as_sf(df_nonzeros,
                           coords = c("Longitude", "Latitude"),
                           crs = 4326,
                           remove = FALSE) %>%
    st_transform(crs = 32717)
  
  # Greedy deduplication: keep as many points as possible >= d apart
  # No sorting by Counts — order doesn't matter, just maximise retained points
  kept_indices <- c()
  
  for (i in 1:nrow(sf_nonzeros)) {
    if (length(kept_indices) == 0) {
      kept_indices <- c(kept_indices, i)
    } else {
      kept_so_far <- sf_nonzeros[kept_indices, ]
      current_point <- sf_nonzeros[i, ]
      distances <- st_distance(current_point, kept_so_far)
      if (all(distances > d)) {
        kept_indices <- c(kept_indices, i)
      }
    }
  }
  
  df_nonzeros_dedup <- sf_nonzeros[kept_indices, ] %>%
    st_drop_geometry()
  
  # Recombine zeros and deduplicated non-zeros
  df_dedup <- bind_rows(df_zeros, df_nonzeros_dedup)
  
  cat("Original points:", nrow(df_data_species), "\n")
  cat("  of which Counts == 0:", nrow(df_zeros), "\n")
  cat("  of which Counts > 0:", nrow(df_nonzeros), "\n")
  cat("Deduplicated non-zero points:", nrow(df_nonzeros_dedup), "\n")
  cat("Final dataset:", nrow(df_dedup), "points\n")
  
  return(df_dedup)
}

# Apply to both species using grid-deduplicated data from previous step
df_data_spider_monkey_dedup2 <- deduplicate_by_distance_v2(df_data_spider_monkey_dedup, d = 500)
df_data_howler_monkey_dedup2 <- deduplicate_by_distance_v2(df_data_howler_monkey_dedup, d = 500)
```

```{r}
df_data_spider_monkey <- df_data_spider_monkey_dedup2
df_data_howler_monkey <- df_data_howler_monkey_dedup2
```


# Data description

## Dataframe structure

```{r dataframe structure, echo = FALSE}
head(df_data)
```

## Number of audio files with at least one detection

```{r, echo = FALSE}
# Number of audio files with at least one detection
cat("Spider Monkey:", sum(df_data_spider_monkey$Counts), "\n")
cat("Howler Monkey:", sum(df_data_howler_monkey$Counts), "\n\n")
```

## Number of study sites with at least one detection

```{r, echo = FALSE}
# Number of audio files with at least one detection
cat("Spider Monkey:", df_data_spider_monkey %>% filter(Occurrence == 1) %>% nrow(), "\n")
cat("Howler Monkey:", df_data_howler_monkey %>% filter(Occurrence == 1) %>% nrow(), "\n")
```

```{r}
# Mean intact forest cover for observations
mean(df_data_spider_monkey$Forest_JRC_2000m[df_data_spider_monkey$Occurrence == 1])
mean(df_data_howler_monkey$Forest_JRC_2000m[df_data_howler_monkey$Occurrence == 1])

# SD intact forest cover for observations
sd(df_data_spider_monkey$Forest_JRC_2000m[df_data_spider_monkey$Occurrence == 1])
sd(df_data_howler_monkey$Forest_JRC_2000m[df_data_howler_monkey$Occurrence == 1])
```

## Correlation between predictors

To reduce multicollinearity among predictors, we generated a correlation matrix including all retained environmental variables. Following Dormann et al. (2013), pairs of predictors with |r| > 0.70 were considered highly correlated and were therefore not included together in the same model.

```{r, echo = FALSE}
# Select only numeric variables from the dataframe
env_vars <- df_data %>% 
  select(where(is.numeric), -Counts, -Occurrence, -Latitude_scaled, -Longitude_scaled)

# Compute correlation matrix
cor_matrix <- cor(env_vars, use = "pairwise.complete.obs")

# Plot
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  diag = FALSE,
  tl.cex = 0.6,
  tl.col = "black",
  tl.srt = 45,
  addCoef.col = "black",
  number.cex = 0.5
)
```

## Map of the study area

```{r}
# Create a bounding square polygon
square_coords <- matrix(c(
  -79.3, 0.36,  # Bottom-left
  -78.896, 0.36,  # Bottom-right
  -78.896, 0.74,  # Top-right
  -79.3, 0.74,  # Top-left
  -79.3, 0.36   # Close the square
), ncol = 2, byrow = TRUE)

study_area <- st_polygon(list(square_coords)) %>%
  st_sfc(crs = 4326) %>%
  st_sf()
```

```{r}
# Load JRC_TMF raster
map_JRCTMF <- rast("C:/Users/nassi/Desktop/Research/MKY_Monkeys/MKY_Data/MKY_Study_sites/Map/JRC_TMF.tif") %>%
 crop(study_area) %>% # Crop to the study area
 mask(study_area)

# As dataframe
df_JRCTMF <- as.data.frame(map_JRCTMF, xy = TRUE, na.rm = TRUE)
```

```{r}
data_plot <- deduplicate_by_grid(df_data) %>%
  distinct(Latitude, Longitude, .keep_all = TRUE)

# Map of species occurrence
ggplot() +
  geom_raster(data = df_JRCTMF, aes(x = x, y = y, fill = as.factor(UndisturbedDegradedForest)),
              alpha = 0.5) +
  scale_fill_manual(
    name = "Land-Use Type",
    values = c("1" = "forestgreen", "2" = "lightgreen", "3" = "steelblue"),
    labels = c("1" = "Intact forest", "2" = "Degraded forest", "3" = "Non-forest")
  ) +
  geom_point(data = data_plot,
    aes(x = Longitude, y = Latitude),
    size = 3,
    alpha = 0.4,
    color = "black"
  ) +
  coord_sf(crs = 4326,expand = FALSE) +
  scale_x_continuous(
  breaks = seq(floor(min(df_data$Longitude)),
                ceiling(max(df_data$Longitude)),
                by = 0.09)
  ) +
  scale_y_continuous(position = "right") +
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  ) +
  labs(
    x = "Longitude", y = "Latitude"
  )
```

# Habitat Suitability Model

We will develop habitat suitability models for spider monkeys and howler monkeys using GAM. First, we will model species occurrence using presence–absence data. Second, we will model vocalization activity using count data, defined as the number of audio files in which each species was detected at each sampling site.

We performed backward selection to identify the most parsimonious models. Model selection was performed as for the occurrence models was based on the Akaike Information Criterion (AIC) (Akaike, 1998; Burnham & Anderson, 2004). Because AIC is not defined for models using a quasi-likelihood distribution, selection of detection count models was instead guided by adjusted R² values and the statistical significance of predictors. 

```{r, include = FALSE}
# Variables
forest_vars <- c("Forest_ESA_500m", "Forest_ESA_2000m", "Forest_JRC_500m", "Forest_JRC_2000m")
integrity_vars <- c("Integrity_500m", "Integrity_2000m")
ndvi_vars <- c("NDVI_500m", "NDVI_2000m")
road_vars <- c("Road_density_500m", "Road_density_2000m")
```

```{r, include = FALSE}
# Helper function to fit models and compute AIC + adjusted R²
fit_gam_binomial <- function(vars, data, response) {
  aic_values <- numeric(length(vars))
  adj_r2_values <- numeric(length(vars))
  names(aic_values) <- vars
  names(adj_r2_values) <- vars
  for (v in vars) {
    f <- as.formula(paste0(response, " ~ ", v))
    model <- gam(f, data = data, family = binomial(link = "log"), select = TRUE)
    sm <- summary(model)
    aic_values[v] <- AIC(model) # extract AIC
    adj_r2_values[v] <- sm$r.sq # extract adjusted R²
  }
  tibble(
    Variable = vars,
    AIC = aic_values,
    Adj_R2 = adj_r2_values
  )
}
```

```{r, include = FALSE}
# Helper function to fit models and compute adjusted R²
fit_gam_quasipoisson <- function(vars, data, response) {
  aic_values <- numeric(length(vars))
  adj_r2_values <- numeric(length(vars))
  names(aic_values) <- vars
  names(adj_r2_values) <- vars
  for (v in vars) {
    f <- as.formula(paste0(response, " ~ ", v))
    model <- gam(f, data = data, family = quasipoisson(link = "log"), select = TRUE)
    sm <- summary(model)
    adj_r2_values[v] <- sm$r.sq # extract adjusted R²
  }
  tibble(
    Variable = vars,
    Adj_R2 = adj_r2_values
  )
}
```

## Brown-headed Spider Monkey

### Hypotheses

- **Forest integrity**  
  Forest integrity is expected to be positively correlated with spider monkey occurrence. Below a certain threshold of forest integrity, spider monkeys are not expected to be detected.

- **Forest cover**  
  Forest cover is expected to be positively correlated with spider monkey occurrence. We expect Forest_JRC to be a better explanatory variable than Forest_ESA, as Forest_JRC quantifies the cover of primary forest only, whereas Forest_ESA includes both primary and disturbed forests. Spider monkeys are expected to avoid disturbed and secondary forests.

- **Canopy height**  
  As a proxy for secondary forest successional stage, canopy height is expected to be positively correlated with spider monkey presence.

- **NDVI**  
  As a proxy for food availability and vegetation productivity, NDVI is expected to be positively correlated with spider monkey presence.

- **Distance to water (D_water)**  
  Distance to water is expected to be negatively associated with spider monkey presence, with higher occurrence closer to water sources.

- **Distance to forest edge (D_edge)**  
  Distance to forest edge is expected to be positively correlated with spider monkey presence. We expect D_edge_JRC to be a better explanatory variable than D_edge_ESA, as spider monkeys are expected to avoid disturbed and secondary forest edges.
  
- **Road density (Road_density)**  
  Road density is expected to be negatively correlated with spider monkey presence. Studies have shown that spider monkeys were sensitive to fragmentation and were unlikely to cross roads, as the canopy cover was not continuous. 
  
### Presence/absence (occurrence)

#### Scale of effect

Because the scale at which landscape variables best predict monkey responses (i.e. the scale of effect) was not known a priori, we calculated FLII, NDVI, and forest cover within circular buffers of 500 m and 2000 m radii around each sampling site. We then evaluated which spatial scale provided the best explanatory power for each species and response variable.

```{r, echo = FALSE}
# Fit each variable group
forest_results <- fit_gam_binomial(forest_vars, df_data_spider_monkey, "Occurrence") %>% mutate(Group = "Forest Cover")
integrity_results <- fit_gam_binomial(integrity_vars, df_data_spider_monkey, "Occurrence") %>% mutate(Group = "Integrity")
ndvi_results <- fit_gam_binomial(ndvi_vars, df_data_spider_monkey, "Occurrence") %>% mutate(Group = "NDVI")
road_results <- fit_gam_binomial(road_vars, df_data_spider_monkey, "Occurrence") %>% mutate(Group = "Road")

# Combine all results
all_results <- bind_rows(forest_results, integrity_results, ndvi_results, road_results) %>%
  arrange(Group, AIC) # Rank by AIC (best = lowest)

# Print ranked results to console
cat("\nModel ranking by AIC (best first):\n")
print(all_results)
```

**Variables retained:** Forest_JRC_500m, Forest_ESA_2000m, Integrity_500m, NDVI_500m, Road_density_2000m.

```{r, include = FALSE}
# Select variables retained
df_data_spider_monkey_occurrence <- df_data_spider_monkey %>% 
  select(
    1:6,
    Elevation, Integrity_500m, Forest_ESA_2000m, Forest_JRC_500m,
    Canopy_height_lang_100m, D_water, NDVI_500m, Slope,
    D_edge_JRC, D_edge_ESA, Road_density_2000m, 
    (ncol(.)-4):ncol(.) # selects the last 3 columns
  )
```

#### Correlation matrix

Forest_ESA_2000m and Integrity_500m are too correlated to be included in the same model. Forest_ESA_2000m and D_edge_JRC are too correlated to be included in the same model. Road_density_2000m and integrity_500m are too correlated to be included together in the same model. We will compute different GAM models to consider these variables independently.

#### GAM with Integrity and D_edge_JRC

```{r, include = FALSE}
# Binary GAM model
gam_spider <- gam(
 Occurrence ~ 
    Forest_JRC_500m +
    Integrity_500m +
    NDVI_500m +
    D_water +
    D_edge_JRC +
    D_edge_ESA +
    Canopy_height_lang_100m +
    Elevation +
   s(Latitude_scaled, Longitude_scaled, bs="tp", k = 10),
 data = df_data_spider_monkey,
 family = binomial(link = "logit"),
 select = TRUE
 )

# Model summary and diagnostics
summary(gam_spider)
cat("AIC:", AIC(gam_spider))
cat("Adj_R²", summary(gam_spider)$r.sq)
```

#### GAM with Forest_ESA and Road_density_2000m

```{r, include = FALSE}
# Binary GAM model
gam_spider <- gam(
 Occurrence ~ 
    Forest_JRC_500m +
    Road_density_2000m +
    Forest_ESA_2000m +
    Slope +
    NDVI_500m +
    D_water +
    D_edge_ESA +
    Canopy_height_lang_100m +
    Elevation +
   s(Latitude_scaled, Longitude_scaled, bs="tp", k = 10),
 data = df_data_spider_monkey,
 family = binomial(link = "logit"),
 select = TRUE
 )

# Model summary and diagnostics
summary(gam_spider)
cat("AIC:", AIC(gam_spider))
cat("Adj_R²", summary(gam_spider)$r.sq)
```

#### Best models after backward selection

##### Partial effects

```{r}
# Best models for spider monkey occurrence
# Spider monkey GAM models within delta AIC <= 2
# Ordered by AIC (lowest first)

# M1: Forest + NDVI (AIC 172.676)
gam_spider_M1 <- gam(
  Occurrence ~
    Forest_JRC_500m +
    NDVI_500m +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),
  data = df_data_spider_monkey,
  family = binomial(link = "logit"),
  select = TRUE
)

# M2: Forest + NDVI + D_water (AIC 172.822)
gam_spider_M2 <- gam(
  Occurrence ~
    Forest_JRC_500m +
    NDVI_500m +
    D_water +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),
  data = df_data_spider_monkey,
  family = binomial(link = "logit"),
  select = TRUE
)

# M3: Forest + Road_density (AIC 173.522)
gam_spider_M3 <- gam(
  Occurrence ~
    Forest_JRC_500m +
    Road_density_2000m +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),
  data = df_data_spider_monkey,
  family = binomial(link = "logit"),
  select = TRUE
)

# M4: Forest + Integrity + NDVI + D_water (AIC 173.551)
gam_spider_M4 <- gam(
  Occurrence ~
    Forest_JRC_500m +
    Integrity_500m +
    NDVI_500m +
    D_water +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),
  data = df_data_spider_monkey,
  family = binomial(link = "logit"),
  select = TRUE
)

# M5: Forest + Road_density + D_water (AIC 173.956)
gam_spider_M5 <- gam(
  Occurrence ~
    Forest_JRC_500m +
    Road_density_2000m +
    D_water +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),
  data = df_data_spider_monkey,
  family = binomial(link = "logit"),
  select = TRUE
)

# M6: Forest + Road_density + NDVI + D_water (AIC 174.354)
gam_spider_M6 <- gam(
  Occurrence ~
    Forest_JRC_500m +
    Road_density_2000m +
    NDVI_500m +
    D_water +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),
  data = df_data_spider_monkey,
  family = binomial(link = "logit"),
  select = TRUE
)

# M7: Forest only (AIC 174.619)
gam_spider_M7 <- gam(
  Occurrence ~
    Forest_JRC_500m +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),
  data = df_data_spider_monkey,
  family = binomial(link = "logit"),
  select = TRUE
)
```

```{r, include = FALSE}
plot_predictions_occurrence <- function(pred_data, var_name, raw_data, x_var) {
  ggplot(pred_data, aes(x = x, y = predicted)) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
                alpha = 0.3, fill = "grey") +
    geom_line(color = "black", linewidth = 2.5) +
    geom_point(data = raw_data,
               aes_string(x = x_var, y = "Occurrence"),
               alpha = 0.3, size = 4, color = "black") +
    labs(x = var_name, y = "Species occurrence") +
    theme_classic() +
    theme(
      aspect.ratio = 1,
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
      axis.title = element_text(size = 20),
      axis.text = element_text(size = 16)
    )
}
```

```{r, echo = FALSE}
# Generate predictions for each variable
pred_forest <- ggpredict(gam_spider_M1, terms = "Forest_JRC_500m [all]")
pred_NDVI <- ggpredict(gam_spider_M1, terms = "NDVI_500m [all]")
pred_D_water <- ggpredict(gam_spider_M2, terms = "D_water [all]")
pred_road <- ggpredict(gam_spider_M3, terms = "Road_density_2000m [0:0.000664 by=0.000001]")

# Create individual plots
plot_predictions_occurrence(pred_forest, "Intact forest cover (%) (500m)", 
                     df_data_spider_monkey, "Forest_JRC_500m")
plot_predictions_occurrence(pred_NDVI, "Normalized Difference Vegetation\nIndex (NDVI) (500m)", 
                     df_data_spider_monkey, "NDVI_500m")
plot_predictions_occurrence(pred_D_water, "Distance to water (m)", 
                     df_data_spider_monkey, "D_water")
plot_predictions_occurrence(pred_road, "Road density (m/m²) (2000m)", 
                     df_data_spider_monkey, "Road_density_2000m")
```

##### Predictions

```{r}
# Store models in a named list
models <- list(
  m1 = gam_spider_M1,
  m2 = gam_spider_M2,
  m3 = gam_spider_M3,
  m4 = gam_spider_M4,
  m5 = gam_spider_M5,
  m6 = gam_spider_M6,
  m7 = gam_spider_M7
)

# Extract AIC
aic_df <- data.frame(
  model = names(models),
  AIC = sapply(models, AIC)
)

# Compute Delta AIC
aic_df$delta_AIC <- aic_df$AIC - min(aic_df$AIC)

# Compute Akaike weights
aic_df$weight <- exp(-0.5 * aic_df$delta_AIC)
aic_df$weight <- aic_df$weight / sum(aic_df$weight)

aic_df

selected_models <- models[aic_df$delta_AIC <= 2]
selected_weights <- aic_df$weight[aic_df$delta_AIC <= 2]
```

```{r}
# Predict for each model
pred_list <- lapply(selected_models, function(m) {
  predict(m, newdata = df_grid, type = "response")
})

# Convert to matrix (columns = models)
pred_matrix <- do.call(cbind, pred_list)
```

```{r}
df_grid <- df_grid %>%
  mutate(
    HSM_spider_occurrence = as.vector(pred_matrix %*% selected_weights)
  )
```

```{r, echo = FALSE}
# Convert df_grid to sf for plotting
df_grid_sf <- st_as_sf(df_grid, coords = c("Longitude", "Latitude"), crs = 4326)

ggplot(df_grid_sf) +
  geom_sf(
    aes(color = HSM_spider_occurrence),
    size = 0.57,
    shape = 15
  ) +
  scale_color_gradientn(
    colours = palette_HSM,
    limits = c(0, 1),
    name = "Habitat\nsuitability\n"
  ) +
  coord_sf(expand = FALSE) +
  scale_x_continuous(
    breaks = seq(
      floor(min(st_coordinates(df_grid_sf)[,1])),
      ceiling(max(st_coordinates(df_grid_sf)[,1])),
      by = 0.09
    )
  ) +
  guides(
    color = guide_colorbar(
      barheight = unit(6, "cm"),
      barwidth  = unit(0.6, "cm"),
      ticks = TRUE,
      frame.colour = "black",
      frame.linewidth = 0.5
    )
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 1
    ),
    legend.key = element_rect(
      colour = "black",
      linewidth = 0.4
    ),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 11),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 11)
  ) +
  labs(
    title = "Predicted Habitat Suitability (occurrence)",
    x = "Longitude",
    y = "Latitude"
  )
```

#### Cross-validation

Model predictive performance was assessed using leave-one-out cross-validation (LOOCV). For each observation, a generalized additive model (GAM) was fitted to the remaining n-1 data points, and the fitted model was then used to predict the excluded observation. This procedure was repeated for all observations to cross-validated predictions for the entire dataset. All models were refitted at each iteration, including re-estimation of smoothing parameters.

```{r, include = FALSE}
# Function to perform LOOCV for any model
loocv_gam <- function(formula, data, family, ...) {
  n <- nrow(data)
  pred <- numeric(n)
  for (i in seq_len(n)) {
    m <- gam(
      formula = formula,
      data    = data[-i, ],
      family  = family,
      ...
    )
    pred[i] <- predict(
      m,
      newdata = data[i, ],
      type = "response"
    )
  }
  return(pred)
}
```

```{r}
loocv_gam_multi <- function(formulas, data, family, weights) {
  
  n_models <- length(formulas)
  n <- nrow(data)
  
  if (is.null(weights)) {
    weights <- rep(1 / n_models, n_models)
  }
  
  pred_mat <- matrix(NA, nrow = n, ncol = n_models)
  
  pb <- txtProgressBar(min = 0, max = n, style = 3)
  
  for (i in seq_len(n)) {
    
    train_data <- data[-i, ]
    test_data  <- data[i, ]
    
    for (j in seq_len(n_models)) {
      
      m <- gam(
        formula = formulas[[j]],
        data    = train_data,
        family  = family,
        select  = TRUE
      )
      
      pred_mat[i, j] <- predict(
        m,
        newdata = test_data,
        type = "response"
      )
    }
    
    setTxtProgressBar(pb, i)
  }
  
  close(pb)
  
  avg_pred <- as.vector(pred_mat %*% weights)
  
  return(avg_pred)
}
```

```{r, echo = FALSE}
# Assess model performance in prediction
formulas <- list(
  # M1: Forest + NDVI (AIC 172.676)
  Occurrence ~
    Forest_JRC_500m +
    NDVI_500m +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),

  # M2: Forest + NDVI + D_water (AIC 172.822)
  Occurrence ~
    Forest_JRC_500m +
    NDVI_500m +
    D_water +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),

  # M3: Forest + Road_density (AIC 173.522)
  Occurrence ~
    Forest_JRC_500m +
    Road_density_2000m +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),

  # M4: Forest + Integrity + NDVI + D_water (AIC 173.551)
  Occurrence ~
    Forest_JRC_500m +
    Integrity_500m +
    NDVI_500m +
    D_water +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),

  # M5: Forest + Road_density + D_water (AIC 173.956)
  Occurrence ~
    Forest_JRC_500m +
    Road_density_2000m +
    D_water +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),

  # M6: Forest + Road_density + NDVI + D_water (AIC 174.354)
  Occurrence ~
    Forest_JRC_500m +
    Road_density_2000m +
    NDVI_500m +
    D_water +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),

  # M7: Forest only (AIC 174.619)
  Occurrence ~
    Forest_JRC_500m +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10)
)

results <- loocv_gam_multi(
  formulas = formulas,
  data = df_data_spider_monkey,
  family = binomial(link = "logit"),
  weights = selected_weights
)
```

```{r}
obs <- df_data_spider_monkey$Occurrence

# AUC
AUC <- auc(obs, results)

# Brier score
Brier <- mean((obs - results)^2)

# Classification accuracy
class_pred <- ifelse(results >= 0.5, 1, 0)
Accuracy <- mean(class_pred == obs)

cat("AUC:", AUC, "\n")
cat("Brier:", Brier, "\n")
cat("Accuracy:", Accuracy)
```

#### Proportion of suitable habitat in each class

```{r}
interval_labels <- c(
  "[0.00, 0.10]",
  "(0.10, 0.20]",
  "(0.20, 0.30]",
  "(0.30, 0.40]",
  "(0.40, 0.50]",
  "(0.50, 0.60]",
  "(0.60, 0.70]",
  "(0.70, 0.80]",
  "(0.80, 0.90]",
  "(0.90, 1.00]"
)

df_grid$HSM_cat10 <- cut(
  df_grid$HSM_spider_occurrence,
  breaks = seq(0, 1, length.out = 11),
  include.lowest = TRUE,
  labels = interval_labels
)

ggplot(df_grid, aes(x = Forest_JRC_500m, fill = HSM_cat10)) +
  geom_density(position = "fill", alpha = 0.9, size = 0.8) +
  scale_fill_manual(values = palette_HSM) +
  coord_sf(expand = FALSE) +
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 11),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 11)
  ) +
  labs(
    fill = "Habitat\nsuitability",
    x = "Intact forest cover (%) (500m)",
    y = "Proportion"
  )
```

```{r}
# Load JRC_TMF raster
map_jrc_tmf <- rast("C:/Users/nassi/Desktop/Research/MKY_Monkeys/MKY_Data/MKY_Study_sites/Map/JRC_TMF.tif") %>%
  crop(study_area) %>% # Crop to the study area
  mask(study_area)

# Convert raster to points (one point per pixel)
pix <- as.points(map_jrc_tmf)       # terra SpatVector of pixel centroids
pix_sf <- st_as_sf(pix)             # convert to sf for nearest-neighbour join


idx <- st_nearest_feature(pix_sf, df_grid_sf)


pix_sf$HSM_spider_occurrence <- df_grid_sf$HSM_spider_occurrence[idx]

summary_table <- pix_sf %>%
  st_drop_geometry() %>%                 # drop geometry to speed up
  group_by(UndisturbedDegradedForest) %>%
  summarise(
    total_pixels = n(),
    pixels_HSM_gt_0.5 = sum(HSM_spider_occurrence > 0.5, na.rm = TRUE),
    proportion = round(pixels_HSM_gt_0.5 / total_pixels, 4)*100
  )
```

### Counts (abundance)

#### Scale of effect

```{r, echo = FALSE}
# Fit each variable group
forest_results <- fit_gam_quasipoisson(forest_vars, df_data_spider_monkey, "Counts") %>% mutate(Group = "Forest Cover")
integrity_results <- fit_gam_quasipoisson(integrity_vars, df_data_spider_monkey, "Counts") %>% mutate(Group = "Integrity")
ndvi_results <- fit_gam_quasipoisson(ndvi_vars, df_data_spider_monkey, "Counts") %>% mutate(Group = "NDVI")
road_results <- fit_gam_quasipoisson(road_vars, df_data_spider_monkey, "Counts") %>% mutate(Group = "Road")

# Combine all results
all_results <- bind_rows(forest_results, integrity_results, ndvi_results, road_results) %>%
  arrange(Group, -Adj_R2)  # Rank by Adj_R2 (best = higher)

# Print ranked results to console
cat("\nModel ranking by Adj_R² (best first):\n")
print(all_results)
```

**Variables retained:** Forest_JRC_500m, Forest_ESA_500m, Integrity_500m, NDVI_500m, Road_density_2000m.

```{r, include = FALSE}
# Select relevant variables
df_data_spider_monkey_counts <- df_data_spider_monkey %>% 
  select(
    1:6,
    Elevation, Integrity_500m, Forest_ESA_500m, Forest_JRC_500m,
    Canopy_height_lang_100m, D_water, NDVI_500m, Slope,
    D_edge_JRC, D_edge_ESA, Road_density_2000m,
    (ncol(.)-4):ncol(.) # selects the last 5 columns
  )
```

#### Correlation matrix

Forest_ESA_500m and Integrity_500m are too correlated to be included in the same model. Forest_ESA_500m and D_edge_JRC are too correlated to be included in the same model. Road_density_2000m and integrity_500m are too correlated to be included together in the same model.We will compute different GAM models to consider these variables independently.

#### GAM with Integrity and D_edge_JRC

```{r}
# GAM with integrity after backward selection
gam_spider <- gam(
  Counts ~ 
    Forest_JRC_500m +
    NDVI_500m +
    Elevation +
   s(Latitude_scaled, Longitude_scaled, bs="tp", k = 10),
  data = df_data_spider_monkey,
  family = quasipoisson(link = "log"),
  select = TRUE
)

# Model summary and diagnostics
summary(gam_spider)
cat("Adj_R²", summary(gam_spider)$r.sq)
```

#### GAM with Forest_ESA and Road_density_2000m

```{r}
# GAM with Forest ESA after backward selection
gam_spider <- gam(
  Counts ~ 
    Forest_JRC_500m +
    NDVI_500m +
    Elevation +
   s(Latitude_scaled, Longitude_scaled, bs="tp", k = 10),
  data = df_data_spider_monkey,
  family = quasipoisson(link = "log"),
  select = TRUE
)

# Model summary and diagnostics
summary(gam_spider)
cat("Adj_R²", summary(gam_spider)$r.sq)
```

#### Best model based on p-values

```{r}
# Best model for spider monkey counts based on p-values
gam_spider_counts_1 <- gam(
 Counts ~ 
    Forest_JRC_500m +
    NDVI_500m +
    Elevation +
   s(Latitude_scaled, Longitude_scaled, bs="tp", k = 10),
 data = df_data_spider_monkey,
  family = quasipoisson(link = "log"),
 select = TRUE
 )
```

##### Predictions

```{r, include = FALSE}
# Helper function: plot function with data points
plot_predictions_counts <- function(pred_data, var_name, raw_data, x_var) {
  ggplot(pred_data, aes(x = x, y = predicted)) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
                alpha = 0.3, fill = "grey") +
    geom_line(color = "black", linewidth = 2.5) +
    geom_point(data = raw_data,
               aes_string(x = x_var, y = "Counts"),
               alpha = 0.3, size = 4, color = "black") +
    labs(x = var_name, y = "Species counts") +
    theme_classic() +
    theme(
      aspect.ratio = 1,
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
      axis.title = element_text(size = 20),
      axis.text = element_text(size = 16)
    )
}
```

```{r, echo = FALSE}
# Generate predictions for each variable
pred_forest <- ggpredict(gam_spider_counts_1, terms = "Forest_JRC_500m [all]")
pred_NDVI <- ggpredict(gam_spider_counts_1, terms = "NDVI_500m [all]")
pred_elevation <- ggpredict(gam_spider_counts_1, terms = "Elevation [all]")

# Create individual plots
plot_predictions_counts(pred_forest, "Primary forest cover (%)", 
                     df_data_spider_monkey, "Forest_JRC_500m")
plot_predictions_counts(pred_NDVI, "Normalized Difference Vegetation Index (NDVI)", 
                     df_data_spider_monkey, "NDVI_500m")
plot_predictions_counts(pred_elevation, "Elevation (m)", 
                     df_data_spider_monkey, "Elevation")
```

```{r, include = FALSE}
# Predict habitat suitability
df_grid <- df_grid %>%
  mutate(
    HSM_spider_counts_1 = predict(
      gam_spider_counts_1,
      newdata = df_grid,
      type = "response"
    ),
    HSM_spider_counts_1_rel =
      (HSM_spider_counts_1 - min(HSM_spider_counts_1)) /
      (max(HSM_spider_counts_1) - min(HSM_spider_counts_1)) # Rescale betzeen 0 and 1
  )
```

```{r, echo = FALSE}
# Convert df_grid to sf for plotting
df_grid_sf <- st_as_sf(df_grid, coords = c("Longitude", "Latitude"), crs = 4326)

ggplot(df_grid_sf) +
  geom_sf(aes(color = HSM_spider_counts_1_rel), size = 0.57, shape = 15) +
  scale_color_gradientn(
    colours = palette_HSM,
    trans = "sqrt",
    limits = c(0, 1),
    name = "Habitat\nsuitability\n"
  ) +
  coord_sf(expand = FALSE) +
  scale_x_continuous(
  breaks = seq(floor(min(st_coordinates(df_grid_sf)[,1])),
                ceiling(max(st_coordinates(df_grid_sf)[,1])),
                by = 0.09)
  ) +
  guides(
    color = guide_colorbar(
      barheight = unit(6, "cm"),
      barwidth  = unit(0.6, "cm"),
      ticks = TRUE,
      frame.colour = "black",
      frame.linewidth = 0.5
    )
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 1
    ),
    legend.key = element_rect(
      colour = "black",
      linewidth = 0.4
    ),
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10)
  ) +
  labs(
    title = "Predicted Habitat Suitability (counts)",
    x = "Longitude", y = "Latitude"
  )
```

#### Best model based on Adj-R²

```{r}
# best model for based on Adj R²
gam_spider_counts_2 <- gam(
 Counts ~ 
    Forest_JRC_500m +
    Integrity_500m +
    NDVI_500m +
    Canopy_height_lang_100m +
    Elevation +
   s(Latitude_scaled, Longitude_scaled, bs="tp", k = 10),
 data = df_data_spider_monkey,
  family = quasipoisson(link = "log"),
 select = TRUE
 )
```

##### Predictions

```{r, echo = FALSE}
# Generate predictions for each variable
pred_forest <- ggpredict(gam_spider_counts_2, terms = "Forest_JRC_500m [all]")
pred_integrity <- ggpredict(gam_spider_counts_2, terms = "Integrity_500m [all]")
pred_NDVI <- ggpredict(gam_spider_counts_2, terms = "NDVI_500m [all]")
pred_elevation <- ggpredict(gam_spider_counts_2, terms = "Elevation [all]")
pred_canopy_height <- ggpredict(gam_spider_counts_2, terms = "Canopy_height_lang_100m [all]")

# Create individual plots
plot_predictions_counts(pred_forest, "Primary forest cover (%)", 
                     df_data_spider_monkey, "Forest_JRC_500m")
plot_predictions_counts(pred_integrity, "Forest Landscape Integrity Index (500m)", 
                     df_data_spider_monkey, "Integrity_500m")
plot_predictions_counts(pred_NDVI, "NDVI (500 m)", 
                     df_data_spider_monkey, "NDVI_500m")
plot_predictions_counts(pred_elevation, "Elevation (m)", 
                     df_data_spider_monkey, "Elevation")
plot_predictions_counts(pred_canopy_height, "Canopy height (m)", 
                     df_data_spider_monkey, "Canopy_height_lang_100m")

```

```{r, include = FALSE}
# Predict habitat suitability
df_grid <- df_grid %>%
  mutate(
    HSM_spider_counts_2 = predict(
      gam_spider_counts_2,
      newdata = df_grid,
      type = "response"
    ),
    HSM_spider_counts_2_rel =
      (HSM_spider_counts_2 - min(HSM_spider_counts_2)) /
      (max(HSM_spider_counts_2) - min(HSM_spider_counts_2)) # Rescale between 0 and 1
  )
```

```{r, echo = FALSE}
# Convert df_grid to sf for plotting
df_grid_sf <- st_as_sf(df_grid, coords = c("Longitude", "Latitude"), crs = 4326)

ggplot(df_grid_sf) +
  geom_sf(aes(color = HSM_spider_counts_2_rel),
          size = 0.57, shape = 15) +
  scale_color_gradientn(
    colours = palette_HSM,
    trans = "sqrt",
    limits = c(0, 1),
    name = "Habitat\nsuitability\n"
  ) +
  coord_sf(expand = FALSE) +
  scale_x_continuous(
  breaks = seq(floor(min(st_coordinates(df_grid_sf)[,1])),
                ceiling(max(st_coordinates(df_grid_sf)[,1])),
                by = 0.09)
  ) +
  guides(
    color = guide_colorbar(
      barheight = unit(6, "cm"),
      barwidth  = unit(0.6, "cm"),
      ticks = TRUE,
      frame.colour = "black",
      frame.linewidth = 0.5
    )
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 1
    ),
    legend.key = element_rect(
      colour = "black",
      linewidth = 0.4
    ),
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10)
  ) +
  labs(
    title = "Predicted Habitat Suitability (counts)",
    x = "Longitude", y = "Latitude"
  )
```

## Ecuadorian Howler Monkey

### Presence/absence (occurrence)

#### Scale of effect

```{r, include = FALSE}
# Fit each variable group
forest_results <- fit_gam_binomial(forest_vars, df_data_howler_monkey, "Occurrence") %>% mutate(Group = "Forest Cover")
integrity_results <- fit_gam_binomial(integrity_vars, df_data_howler_monkey, "Occurrence") %>% mutate(Group = "Integrity")
ndvi_results <- fit_gam_binomial(ndvi_vars, df_data_howler_monkey, "Occurrence") %>% mutate(Group = "NDVI")
road_results <- fit_gam_binomial(road_vars, df_data_howler_monkey, "Occurrence") %>% mutate(Group = "Road")

# Combine all results
all_results <- bind_rows(forest_results, integrity_results, ndvi_results, road_results) %>%
  arrange(Group, AIC)  # Rank by AIC (best = lowest)

# Print ranked results to console
cat("\nModel ranking by AIC (best first):\n")
print(all_results)
```

**Variables retained:** Forest_JRC_500m, Forest_ESA_2000m, Integrity_2000m, NDVI_500m,Road_density_500m.

```{r, include = FALSE}
# Select relevant variables
df_data_howler_monkey_occurrence <- df_data_howler_monkey %>% 
  select(
    1:6,
    Elevation, Integrity_2000m, Forest_ESA_2000m, Forest_JRC_500m,
    Canopy_height_lang_100m, D_water, NDVI_500m, Slope,
    D_edge_JRC, D_edge_ESA, Road_density_2000m,
    (ncol(.)-4):ncol(.) # selects the last 5 columns
  )
```

#### Correlation matrix

Integrity_2000m and Forest_ESA_2000m are highly correlated. D_edge_JRC and Forest_ESA_2000m are highly correlated. We will compute different GAM models to consider these variables independently.

#### GAM with integrity and D_edge_JRC

```{r, include = FALSE}
# Binary GAM model
gam_howler <- gam(
 Occurrence ~ 
    Forest_JRC_500m +
    Integrity_2000m +
    NDVI_500m +
    Elevation,
 data = df_data_howler_monkey,
 family = binomial(link = "logit"),
 select = TRUE
 )

# Model summary and diagnostics
summary(gam_howler)
cat("AIC:", AIC(gam_howler), "// ")
cat("Adj_R²", summary(gam_howler)$r.sq)
```

#### GAM with Forest_ESA

```{r, include = FALSE}
# Binary GAM model
gam_howler <- gam(
 Occurrence ~ 
    Forest_JRC_500m +
    Integrity_2000m +
    NDVI_500m +
    Elevation,
 data = df_data_howler_monkey,
 family = binomial(link = "logit"),
 select = TRUE
 )

# Model summary and diagnostics
summary(gam_howler)
cat("AIC:", AIC(gam_howler), "// ")
cat("Adj_R²", summary(gam_howler)$r.sq)
```

#### Best models

##### Partial effects

```{r}
# Best models for spider monkey occurrence
# Howler monkey GAM models within delta AIC <= 2
# Ordered by AIC (lowest first)

# M1: Forest + Integrity + NDVI + Elevation (AIC 85.275)
gam_howler_M1 <- gam(
  Occurrence ~
    Forest_JRC_500m +
    Integrity_2000m +
    NDVI_500m +
    Elevation,
  data = df_data_howler_monkey,
  family = binomial(link = "logit"),
  select = TRUE
)

# M2: Forest + Integrity + NDVI + D_edge + Canopy + Elevation + spatial smooth (AIC 85.308)
gam_howler_M2 <- gam(
  Occurrence ~
    Forest_JRC_500m +
    Integrity_2000m +
    NDVI_500m +
    D_edge_JRC +
    Canopy_height_lang_100m +
    Elevation +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),
  data = df_data_howler_monkey,
  family = binomial(link = "logit"),
  select = TRUE
)

# M3: Forest + Integrity + NDVI + D_edge + Canopy + Elevation (smooth penalised to zero) (AIC 85.308)
gam_howler_M3 <- gam(
  Occurrence ~
    Forest_JRC_500m +
    Integrity_2000m +
    NDVI_500m +
    D_edge_JRC +
    Canopy_height_lang_100m +
    Elevation,
  data = df_data_howler_monkey,
  family = binomial(link = "logit"),
  select = TRUE
)

# M4: Forest + Integrity + NDVI + Canopy + Elevation (AIC 85.685)
gam_howler_M4 <- gam(
  Occurrence ~
    Forest_JRC_500m +
    Integrity_2000m +
    NDVI_500m +
    Canopy_height_lang_100m +
    Elevation +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),
  data = df_data_howler_monkey,
  family = binomial(link = "logit"),
  select = TRUE
)

# M5: Forest + Road + Integrity + NDVI + D_edge + Canopy + Elevation + spatial smooth (AIC 86.745)
gam_howler_M5 <- gam(
  Occurrence ~
    Forest_JRC_500m +
    Road_density_2000m +
    Integrity_2000m +
    NDVI_500m +
    D_edge_JRC +
    Canopy_height_lang_100m +
    Elevation +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),
  data = df_data_howler_monkey,
  family = binomial(link = "logit"),
  select = TRUE
)
```

```{r, echo = FALSE}
# Generate predictions for each variable
pred_forest <- ggpredict(gam_howler_M1, terms = "Forest_JRC_500m [all]")
pred_NDVI <- ggpredict(gam_howler_M1, terms = "NDVI_500m [all]")
pred_integrity_2000 <- ggpredict(gam_howler_M1, terms = "Integrity_2000m [all]")
pred_elevation <- ggpredict(gam_howler_M1, terms = "Elevation [all]")

# Create individual plots
plot_predictions_occurrence(pred_forest, "Intact forest cover (%) (500m)", 
                     df_data_howler_monkey, "Forest_JRC_500m")
plot_predictions_occurrence(pred_NDVI, "Normalized Difference Vegetation\nIndex (NDVI) (500m)", 
                     df_data_howler_monkey, "NDVI_500m")
plot_predictions_occurrence(pred_integrity_2000, "Forest Landscape Integrity\nIndex (FLII) (2000m)", 
                     df_data_howler_monkey, "Integrity_2000m")
plot_predictions_occurrence(pred_elevation, "Elevation (m)", 
                     df_data_howler_monkey, "Elevation")
```

##### Predictions

```{r}
# Store models in a named list
models <- list(
  m1 = gam_howler_M1,
  m2 = gam_howler_M2,
  m3 = gam_howler_M3,
  m4 = gam_howler_M4,
  m5 = gam_howler_M5
)

# Extract AIC
aic_df <- data.frame(
  model = names(models),
  AIC = sapply(models, AIC)
)

# Compute Delta AIC
aic_df$delta_AIC <- aic_df$AIC - min(aic_df$AIC)

# Compute Akaike weights
aic_df$weight <- exp(-0.5 * aic_df$delta_AIC)
aic_df$weight <- aic_df$weight / sum(aic_df$weight)

aic_df

selected_models <- models[aic_df$delta_AIC <= 2]
selected_weights <- aic_df$weight[aic_df$delta_AIC <= 2]
```

```{r}
# Predict for each model
pred_list <- lapply(selected_models, function(m) {
  predict(m, newdata = df_grid, type = "response")
})

# Convert to matrix (columns = models)
pred_matrix <- do.call(cbind, pred_list)
```

```{r}
df_grid <- df_grid %>%
  mutate(
    HSM_howler_occurrence = as.vector(pred_matrix %*% selected_weights)
  )
```

```{r, echo = FALSE}
# Convert df_grid to sf for plotting
df_grid_sf <- st_as_sf(df_grid, coords = c("Longitude", "Latitude"), crs = 4326)

ggplot(df_grid_sf) +
  geom_sf(aes(color = HSM_howler_occurrence), size = 0.57, shape = 15) +
  scale_color_gradientn(
    colours = palette_HSM,
    limits = c(0, 1),
    name = "Habitat\nsuitability\n"
  ) +
  coord_sf(expand = FALSE) +
  scale_x_continuous(
  breaks = seq(floor(min(st_coordinates(df_grid_sf)[,1])),
                ceiling(max(st_coordinates(df_grid_sf)[,1])),
                by = 0.09)
  ) +
  guides(
    color = guide_colorbar(
      barheight = unit(6, "cm"),
      barwidth  = unit(0.6, "cm"),
      ticks = TRUE,
      frame.colour = "black",
      frame.linewidth = 0.5
    )
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 1
    ),
    legend.key = element_rect(
      colour = "black",
      linewidth = 0.4
    ),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 11),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 11)
  ) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  ) +
  labs(
    title = "Predicted Habitat Suitability (occurrence)",
    x = "Longitude", y = "Latitude"
  )
```

#### Cross-validation

```{r, echo = FALSE}
# Assess model performance in prediction
formulas <- list(
  # M1: Forest + Integrity + NDVI + Elevation (AIC 85.275)
  Occurrence ~
    Forest_JRC_500m +
    Integrity_2000m +
    NDVI_500m +
    Elevation,

  # M2: Forest + Integrity + NDVI + D_edge + Canopy + Elevation + spatial smooth (AIC 85.308)
  Occurrence ~
    Forest_JRC_500m +
    Integrity_2000m +
    NDVI_500m +
    D_edge_JRC +
    Canopy_height_lang_100m +
    Elevation +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),

  # M3: Forest + Integrity + NDVI + D_edge + Canopy + Elevation (smooth penalised to zero) (AIC 85.308)
  Occurrence ~
    Forest_JRC_500m +
    Integrity_2000m +
    NDVI_500m +
    D_edge_JRC +
    Canopy_height_lang_100m +
    Elevation,

  # M4: Forest + Integrity + NDVI + Canopy + Elevation (AIC 85.685)
  Occurrence ~
    Forest_JRC_500m +
    Integrity_2000m +
    NDVI_500m +
    Canopy_height_lang_100m +
    Elevation +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10),

  # M5: Forest + Road + Integrity + NDVI + D_edge + Canopy + Elevation + spatial smooth (AIC 86.745)
  Occurrence ~
    Forest_JRC_500m +
    Road_density_2000m +
    Integrity_2000m +
    NDVI_500m +
    D_edge_JRC +
    Canopy_height_lang_100m +
    Elevation +
    s(Latitude_scaled, Longitude_scaled, bs = "tp", k = 10)
)
results <- loocv_gam_multi(
  formulas = formulas,
  data = df_data_howler_monkey,
  family = binomial(link = "logit"),
  weights = selected_weights
)
```

```{r}
obs <- df_data_howler_monkey$Occurrence

# AUC
AUC <- auc(obs, results)

# Brier score
Brier <- mean((obs - results)^2)

# Classification accuracy
class_pred <- ifelse(results >= 0.5, 1, 0)
Accuracy <- mean(class_pred == obs)

cat("AUC:", AUC, "\n")
cat("Brier:", Brier, "\n")
cat("Accuracy:", Accuracy)
```

#### Proportion of suitable habitat in each class

```{r}
df_grid$HSM_cat10 <- cut(
  df_grid$HSM_howler_occurrence,
  breaks = seq(0, 1, length.out = 11),
  include.lowest = TRUE,
  labels = interval_labels
)

ggplot(df_grid, aes(x = Forest_JRC_500m, fill = HSM_cat10)) +
  geom_density(position = "fill", alpha = 0.9, size = 0.8) +
  scale_fill_manual(values = palette_HSM) +
  coord_sf(expand = FALSE) +
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 11),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 11)
  ) +
  labs(
    fill = "Habitat\nsuitability",
    x = "Intact forest cover (%) (500m)",
    y = "Proportion"
  )
```

```{r}
pix_sf$HSM_howler_occurrence <- df_grid_sf$HSM_howler_occurrence[idx]

summary_table <- pix_sf %>%
  st_drop_geometry() %>%                 # drop geometry to speed up
  group_by(UndisturbedDegradedForest) %>%
  summarise(
    total_pixels = n(),
    pixels_HSM_gt_0.5 = sum(HSM_howler_occurrence > 0.5, na.rm = TRUE),
    proportion = round(pixels_HSM_gt_0.5 / total_pixels, 4)*100
  )
```

### Counts

#### Scale of effect

```{r, include = FALSE}
# Fit each variable group
forest_results <- fit_gam_quasipoisson(forest_vars, df_data_howler_monkey, "Counts") %>% mutate(Group = "Forest Cover")
integrity_results <- fit_gam_quasipoisson(integrity_vars, df_data_howler_monkey, "Counts") %>% mutate(Group = "Integrity")
ndvi_results <- fit_gam_quasipoisson(ndvi_vars, df_data_howler_monkey, "Counts") %>% mutate(Group = "NDVI")
road_results <- fit_gam_quasipoisson(road_vars, df_data_howler_monkey, "Counts") %>% mutate(Group = "Road")

# Combine all results
all_results <- bind_rows(forest_results, integrity_results, ndvi_results, road_results) %>%
  arrange(Group, -Adj_R2)  # Rank by Adj_R2 (best = higher)

# Print ranked results to console
cat("\nModel ranking by Adj_R² (best first):\n")
print(all_results)
```

```{r, echo = FALSE}
# Select relevant variables
df_data_howler_monkey_counts <- df_data_howler_monkey %>% 
  select(
    1:6,
    Elevation, Integrity_500m, Forest_ESA_500m, Forest_JRC_2000m,
    Canopy_height_lang_100m, D_water, NDVI_500m, Slope,
    D_edge_JRC, D_edge_ESA, Road_density_500m,
    (ncol(.)-4):ncol(.) # selects the last 5 columns
  )
```

**Variables retained:** Forest_JRC_2000m, Forest_ESA_500m, Integrity_500m, NDVI_500m, Road_density_500m.

#### Correlation matrix

#### GAM with Integrity and D_edge_JRC

```{r, include = FALSE}
# GAM
gam_howler <- gam(
  Counts ~ 
    Forest_JRC_2000m +
    Integrity_500m +
    Road_density_500m +
    Slope +
    NDVI_500m +
    D_water +
    D_edge_JRC +
    D_edge_ESA +
    Canopy_height_lang_100m +
    Elevation +
   s(Latitude_scaled, Longitude_scaled, bs="tp", k = 5),
  data = df_data_howler_monkey,
  family = quasipoisson(link = "log"),
  select = TRUE
)

gam_howler <- gam(
  Counts ~ 
    Forest_JRC_2000m +
    Integrity_500m +
    NDVI_500m +
    D_water +
    Elevation +
   s(Latitude_scaled, Longitude_scaled, bs="tp", k = 5),
  data = df_data_howler_monkey,
  family = quasipoisson(link = "log"),
  select = TRUE
)

# Model summary and diagnostics
summary(gam_howler)
cat("Adj_R²", summary(gam_howler)$r.sq)
```

#### GAM with Forest_ESA

```{r, include = FALSE}
# GAM
gam_howler <- gam(
  Counts ~ 
    Forest_JRC_2000m +
    Road_density_500m +
    Forest_ESA_500m +
    Slope +
    NDVI_500m +
    D_water +
    D_edge_ESA +
    Canopy_height_lang_100m +
    Elevation +
   s(Latitude_scaled, Longitude_scaled, bs="tp", k = 5),
  data = df_data_howler_monkey,
  family = quasipoisson(link = "log"),
  select = TRUE
)

# Model summary and diagnostics
summary(gam_howler)
cat("Adj_R²", summary(gam_howler)$r.sq)
```

#### Best model based on p-values

```{r}
# Best model for spider monkey counts based on p-values
gam_howler_counts_1 <- gam(
  Counts ~ 
    Integrity_500m +
    NDVI_500m +
    D_water +
    Elevation,
 data = df_data_howler_monkey,
  family = quasipoisson(link = "log"),
 select = TRUE
 )

# Model summary and diagnostics
summary(gam_howler_counts_1)
cat("Adj_R²", summary(gam_howler_counts_1)$r.sq)
```

##### Predictions

```{r}
# Generate predictions for each variable
pred_integrity <- ggpredict(gam_howler_counts_1, terms = "Integrity_500m [all]")
pred_NDVI <- ggpredict(gam_howler_counts_1, terms = "NDVI_500m [all]")
pred_D_water <- ggpredict(gam_howler_counts_1, terms = "D_water [all]")
pred_elevation <- ggpredict(gam_howler_counts_1, terms = "Elevation [all]")

# Create individual plots
plot_predictions_counts(pred_integrity, "Forest Landscape Integrity Index (FLII)", 
                     df_data_howler_monkey, "Integrity_500m")
plot_predictions_counts(pred_NDVI, "Normalized Difference Vegetation Index (NDVI)", 
                     df_data_howler_monkey, "NDVI_500m")
plot_predictions_counts(pred_D_water, "Distance to water (m)", 
                     df_data_howler_monkey, "D_water")
plot_predictions_counts(pred_elevation, "Elevation (m)", 
                     df_data_howler_monkey, "Elevation")
```

```{r, include = FALSE}
# Predict habitat suitability
df_grid <- df_grid %>%
  mutate(
    HSM_howler_counts_1 = predict(
      gam_howler_counts_1,
      newdata = df_grid,
      type = "response"
    ),
    HSM_howler_counts_1_rel =
      (HSM_howler_counts_1 - min(HSM_howler_counts_1)) /
      (max(HSM_howler_counts_1) - min(HSM_howler_counts_1)) # Rescale betzeen 0 and 1
  )
```

```{r, echo = FALSE}
# Convert df_grid to sf for plotting
df_grid_sf <- st_as_sf(df_grid, coords = c("Longitude", "Latitude"), crs = 4326)

ggplot(df_grid_sf) +
  geom_sf(aes(color = HSM_howler_counts_1_rel), size = 0.57, shape = 15) +
  scale_color_gradientn(
    colours = palette_HSM,
    trans = "sqrt",
    limits = c(0, 1),
    name = "Habitat\nsuitability\n"
  ) +
  coord_sf(expand = FALSE) +
  scale_x_continuous(
  breaks = seq(floor(min(st_coordinates(df_grid_sf)[,1])),
                ceiling(max(st_coordinates(df_grid_sf)[,1])),
                by = 0.09)
  ) +
  guides(
    color = guide_colorbar(
      barheight = unit(6, "cm"),
      barwidth  = unit(0.6, "cm"),
      ticks = TRUE,
      frame.colour = "black",
      frame.linewidth = 0.5
    )
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 1
    ),
    legend.key = element_rect(
      colour = "black",
      linewidth = 0.4
    ),
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10)
  ) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  ) +
  labs(
    title = "Predicted Habitat Suitability (counts)",
    x = "Longitude", y = "Latitude"
  )
```
#### Best model based on Adj-R²

```{r}
# Best model for spider monkey counts based on adj R²
gam_howler_counts_2 <- gam(
  Counts ~ 
    Forest_JRC_2000m +
    Integrity_500m +
    NDVI_500m +
    D_water +
    Elevation +
   s(Latitude_scaled, Longitude_scaled, bs="tp", k = 5),
 data = df_data_howler_monkey,
  family = quasipoisson(link = "log"),
 select = TRUE
 )
```

##### Predictions

```{r, echo = FALSE}
# Generate predictions for each variable
pred_forest <- ggpredict(gam_howler_counts_2, terms = "Forest_JRC_2000m [all]")
pred_integrity <- ggpredict(gam_howler_counts_2, terms = "Integrity_500m [all]")
pred_NDVI <- ggpredict(gam_howler_counts_2, terms = "NDVI_500m [all]")
pred_D_water <- ggpredict(gam_howler_counts_2, terms = "D_water [all]")
pred_elevation <- ggpredict(gam_howler_counts_2, terms = "Elevation [all]")

# Create individual plots
plot_predictions_counts(pred_forest, "Forest cover JRC (2000m)", df_data_howler_monkey, "Forest_JRC_2000m")
plot_predictions_counts(pred_integrity, "Forest Landscape Integrity Index (FLII)", df_data_howler_monkey, "Integrity_500m")
plot_predictions_counts(pred_NDVI, "Normalized Difference Vegetation Index (NDVI)", df_data_howler_monkey, "NDVI_500m")
plot_predictions_counts(pred_D_water, "Distance to water (m)", df_data_howler_monkey, "D_water")
plot_predictions_counts(pred_elevation, "Elevation (m)", df_data_howler_monkey, "Elevation")
```

```{r, include = FALSE}
# Predict habitat suitability
df_grid <- df_grid %>%
  mutate(
    HSM_howler_counts_2 = predict(
      gam_howler_counts_2,
      newdata = df_grid,
      type = "response"
    ),
    HSM_howler_counts_2_rel =
      (HSM_howler_counts_2 - min(HSM_howler_counts_2)) /
      (max(HSM_howler_counts_2) - min(HSM_howler_counts_2)) # Rescale betzeen 0 and 1
  )
```

```{r, echo = FALSE}
# Convert df_grid to sf for plotting
df_grid_sf <- st_as_sf(df_grid, coords = c("Longitude", "Latitude"), crs = 4326)

ggplot(df_grid_sf) +
  geom_sf(aes(color = HSM_howler_counts_2_rel), size = 0.57, shape = 15) +
  scale_color_gradientn(
    colours = palette_HSM,
    trans = "sqrt",
    limits = c(0, 1),
    name = "Habitat\nsuitability\n"
  ) +
  coord_sf(expand = FALSE) +
  scale_x_continuous(
  breaks = seq(floor(min(st_coordinates(df_grid_sf)[,1])),
                ceiling(max(st_coordinates(df_grid_sf)[,1])),
                by = 0.09)
  ) +
  guides(
    color = guide_colorbar(
      barheight = unit(6, "cm"),
      barwidth  = unit(0.6, "cm"),
      ticks = TRUE,
      frame.colour = "black",
      frame.linewidth = 0.5
    )
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 1
    ),
    legend.key = element_rect(
      colour = "black",
      linewidth = 0.4
    ),
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10)
  ) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  ) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  ) +
  labs(
    title = "Predicted Habitat Suitability (counts)",
    x = "Longitude", y = "Latitude"
  )
```

```{r}

```


